
\chapter{Acknowledgments} 

This work would not have happened without contributions from many colleagues over the past decade.  We list some of these contributions, and apologize to those colleagues who we forget to mention.  Specifically, we thank Hans Skaug for discussions regarding the Laplace approximation, how to include the SPDE method in TMB, and the implementation and performance of restricted maximum likelihood. We also thank Steve Munch for discussing phylogenetic comparative methods and empirical dynamical models. Thanks to Eli Holmes, Eric Ward, and Mark Scheuerell for past discussions regarding Gompertz density dependence, Jay Ver Hoef for discussing simultaneous autoregressive processes, and Yumi Arimitsu and Bill Sydeman for discussing climate velocity.  Finally, we thank Sergey Feldman for discussions regarding causal analysis and do-calculus, and Julie Nielsen, Kevin Siwicke, and Devin Johnson for discussing the hidden Markov model analysis of the archival tagging data.  

We also feel lucky to include data that were collected and compiled by many different teams worldwide.  We thank Carey Kuhn and Jeremy Stirling for providing data for the northern fur seal case study (as well as discussion of results), and Susanne McDermott, Julie Nielsen, and many team members who collected and provided the Pacific cod archival tag data set.  We thank Richard Condit, Stephen Hubbell, and many others collecting, distributing, and permission to use Barro Colorado data, and the huge number of contributors to the Breeding Bird survey.  We thank the Groundfish Assessment Program at the Alaska Fisheries Science Center for conducting the Bering Sea shelf bottom trawl survey annually since 1982, and the SEAMAP Groundfish Trawl Survey, NMFS Pelagic Acoustic Trawl Survey, and NMFS Red Snapper/Shark Bottom Longline Surveys for collecting the red snapper case-study data (as well as Arnaud Grüss for compiling these data). Finally, thanks to Kevin Siwicke for guidance in accessing the sea ice concentration data, and Connie Okasaki for guidance in accessing the ozone concentration data.  

Finally, we're grateful for ideas and editorial contributions from Baptiste Alglave, Cheryl Barnes, Jie Cao, Matt Cheng, Curry Cunningham, Tim Essington, Devin Johnson, Sami Kivelä, Cole Monnahan, Julie Nielsen, and Sam Urmy.  Remaining errors are obviously our own.  We also thank Anders Nielsen, Casper Berg, Tim Essington, Devin Johnson, and many others for encouragement to write this book, and Nobel Hendrix for co-teaching the initial version of a class that inspired this book.  Finally, we thank countless co-authors, reviewers, journal editors, R-package developers, and colleagues for guiding our research regarding spatio-temporal models over the past decade. 
